A Dataset of Scratch Programs:\\Scraped, Shaped and Scored

### 2017-03-16 MSR-Scratch.tex
Shortened paper and references

The coding elements in the Scratch environment are the \textbf{\texttt{blocks}}, which can receive input parameters.
In the example in the figure, block `go to x:0 y:0' receives as input two numerical parameters.
Custom clock invocations are stored in this table just like regular blocks.
To cater for the varying number of input parameters that those blocks can have, this table can store up to the maximum number of input parameters found in the dataset, which is 24.
The ordering of the blocks in the scripts can be retrieved using the \texttt{block-rank} field, which represents a depth-first ranking of the blocks.
In the script of the `backflip' custom block definition in the example, the ranking of the blocks would be `define backflip', `set counter to', `+', `counter', followed by the rest of the blocks.
The clocks can be of different shapes and categories.
In table \textbf{\texttt{blockTypes}} we have stored the encoding of all types of Scratch blocks and of the blocks from Scratch extensions.

@@

The majority of the projects have been viewed only once, and they have not been favorited or remixed.
Around 10,000 of the total projects have been remixed at least once, while the dataset includes cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.

Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of development of computational thinking skills. The table highlights the existence of surprisingly complex projects, with 525 sprites, 3,000 scripts and 34,000 blocks.

@@

The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education is receiving significant research attention.
A number of studies have been carried out during the last years on understanding of the programming concepts and the programming practices of novice programmers in those environments, on the programming skills they develop, and on the quality of their programs.

@@

Another promising research direction relates to the \textit{quality of the programming artifacts developed in the Scratch environment}.
The dataset has already been used for quality assessment, for the identification of programming smells~\cite{Aivaloglou_2016}, and for exploring indications of harmful programming habits~\cite{Robles_2017}.
Existing works in this direction include a study in a classroom setting~\cite{Meerbaum_habits_2011}, which highlighted two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.
The authors connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops.
Related to smell detection are the Scratch automated quality analysis tools Hairball~\cite{boe_hairball:_2013}, a lint-like static analysis tool for Scratch that can detect initialization problems and unmatched broadcast and receive blocks, and Dr. Scratch~\cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts.

@@

Enabling research towards this direction requires extending the dataset with the complete set of projects for the included users, along with their creation dates.
This would enable examining how the programming skills and mastery of the users evolved through time, possibly through remixing the projects of other users.
An even richer extension of the dataset would be with, possibly daily, snapshots of the included projects that are still in development.
This would enable reconstructing some project versioning information, not available from the Scratch web interface, and thus examining how the projects are developed and evolve over time.

>>>>

The coding elements in the Scratch environment are the \textbf{\texttt{blocks}}, which can receive input parameters, like 
block `go to x:0 y:0' in the figure.
Custom clock invocations are stored in this table just like regular blocks.
To cater for the varying number of input parameters that those blocks can have, this table can store up to the maximum number of input parameters found in the dataset, which is 24.
The ordering of the blocks in the scripts can be retrieved using the \texttt{block-rank} field, which represents a depth-first ranking of the blocks.
In the script of the `backflip' custom block definition in the example, the ranking of the blocks would be `define backflip', `set counter to', `+', `counter', followed by the rest of the blocks.
The clocks can be of different shapes and categories.
In table \textbf{\texttt{blockTypes}} we have stored the encoding of all types of Scratch blocks and of the blocks from Scratch extensions.

@@

The majority of the projects have been viewed only once, and they have not been favorited or remixed.
Around 10,000 of the total projects have been remixed at least once, while there are cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.

Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of computational thinking skills development. The table highlights the existence of surprisingly complex projects, with 525 sprites and over 34,000 blocks.

@@

The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education is receiving significant research attention.
A number of studies have been carried out during the last years on understanding the programming practices of learners in those environments, on the programming skills they develop, and on the quality of their programs.

@@

Another promising research direction relates to the \textit{quality of the programming artifacts developed in the Scratch environment}.
The dataset has already been used for quality assessment, for the identification of programming smells~\cite{Aivaloglou_2016}, and for exploring indications of harmful programming habits~\cite{Robles_2017}.
Existing works in this direction include a study in a classroom setting~\cite{Meerbaum_habits_2011}, which highlighted two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.
The authors connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops.
Related to smell detection are the Scratch automated quality analysis tools Hairball~\cite{boe_hairball:_2013}, a static analysis tool that can detect initialization problems and unmatched broadcast and receive blocks, and Dr. Scratch~\cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits: not changing the default object names and duplicating scripts.

@@

Research towards this direction requires extending the dataset with the complete set of projects for the included users, along with their creation dates.
This would enable examining how their programming skills and mastery evolved through time, possibly through remixing the projects of others.
An even richer extension of the dataset would be with periodical snapshots of the included projects that are still in development.
This would enable reconstructing project versioning information, not available from the Scratch interface, and thus examining how the projects are developed and evolve over time.
